---
title: "Import movement dataset for the Freeze-thaw project"
author: "Ivan Calandra"
date: "`r format(Sys.time(), usetz = TRUE)`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cerulean
    highlight: pygments
    number_sections: true
  github_document: 
    toc: true
    toc_depth: 3
    html_preview: false
bibliography: FT_1_Movement-import.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = rprojroot::find_rstudio_root_file()) })
---


```{r Knitr Options, include = FALSE}
knitr::opts_chunk$set(comment = NA, message = FALSE, indent = "", error = TRUE)
```


---


# Goal of the script
This script formats the movement data.
The script will:

1. Read in the original file  
2. Format the data  
3. Write XLSX file and save R object ready for further analysis in R  

```{r}
dir_in  <- "analysis/raw_data"
dir_out <- "analysis/derived_data"
```

Raw data must be located in "`r paste0("./", dir_in)`".  
Formatted data will be saved in "`r paste0("./", dir_out)`".

The knit directory for this script is the project directory. This is important to specify the path for input and output.  
However, the HTML, MD and BIB files resulting from the rendering of this Rmd file will be located in the same folder as the Rmd file (i.e. "`./analysis/scripts`").

---


# Load packages
```{r}
library(grateful)
library(knitr)
library(R.utils)
library(rmarkdown)
library(tidyverse)
library(writexl)
```


---


# Read and format data
## Read in CSV file
```{r}
mov_file <- list.files(dir_in, pattern = ".*movement.*\\.csv$", full.names = TRUE) 
mov <- read.csv(mov_file, header = TRUE, fileEncoding = 'latin1')
```

The following raw data file has been read: "`r mov_file`"


## Calculate new columns
Create new columns based on combinations of existing columns.  
All values are in cm, except for columns `Freeze-thaw_cycles` and `NO_tube`.

```{r}
mov_all <- mutate(mov,
                  Original_depth_top_point1 = Tube_height - Original_depth_bottom_point1,
                  Original_depth_top_point2 = Tube_height - Original_depth_bottom_point2,
                  Movement_point1 = Final_depth_top_point1 - Original_depth_top_point1,
                  Movement_point2 = Final_depth_top_point2 - Original_depth_top_point2,
                  Abs_movement_point1 = abs(Movement_point1),
                  Abs_movement_point2 = abs(Movement_point2),
                  Average_movement = (Movement_point1 + Movement_point2)/2,
                  Average_movement_CONV = -Average_movement,
                  Total_abs_movement = Abs_movement_point1 + Abs_movement_point2) 
```


## Extract sediment info from column Specimen
```{r}
                    # Split column at every "_"
mov_all$Sediment <- strsplit(mov_all$Specimen, "_") %>% 
  
                    # Keep second value
                    sapply(., "[[", 2) %>% 
  
                    # Replace "fs" and "cs"
                    gsub("fs", "Fine sand", .) %>% 
                    gsub("cs", "Coarse sand", .) %>% 
  
                    # Capitalize first word
                    str_to_sentence()
```


## Keep only SampleID in the column Specimen
```{r}
mov_all$Specimen <- strsplit(mov_all$Specimen, "_") %>% 
                    sapply(., "[[", 1)
```


## Re-order columns and add unit as comment
```{r}
mov_final <- select(mov_all,
                    Specimen, Sediment, Freeze.thaw_cycles, 
                    Original_depth_bottom_point1, Original_depth_bottom_point2,
                    Original_depth_top_point1, Original_depth_top_point2,
                    Final_depth_top_point1, Final_depth_top_point2,
                    Movement_point1:Total_abs_movement,
                    Lithic_rotation, Lithic_tilting)
comment(mov_final) <- "cm"
```

Type `comment(mov_final)` to check the unit of the values.


## Check the result
```{r}
str(mov_final)
head(mov_final)
```


---


# Save data
## As XLSX
```{r}
write_xlsx(list("data" = mov_final, "unit" = data.frame(Unit = paste("All values are in", comment(mov_final)))), 
           path = paste0(dir_out, "/FT_movement_formatted-data.xlsx"))
```


## As Rbin
```{r}
saveObject(mov_final, file = paste0(dir_out, "/FT_movement_formatted-data.Rbin"))
```

Rbin files (e.g. `FT_movement_formatted-data.Rbin`) can be easily read into an R object (e.g. `rbin_data`) using the following code:
```{r, eval = FALSE}
library(R.utils)
rbin_data <- loadObject("FT_movement_formatted-data")
```


---


# sessionInfo()

```{r}
sessionInfo()
```


---


# Cite R packages used

```{r, echo = FALSE}
pkgs_cite <- cite_packages(pkgs = "Session", omit = NULL, output = "table", include.RStudio = TRUE, 
                           out.dir = "analysis/scripts", bib.file = "FT_1_Movement-import")
knitr::kable(pkgs_cite)
```


## References

