---
title: "Import dataset from the surface texture analysis for the Freeze-thaw project"
author: "Ivan Calandra"
date: "`r format(Sys.time(), usetz = TRUE)`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cerulean
    highlight: pygments
    number_sections: true
  github_document: 
    toc: true
    toc_depth: 3
    html_preview: false
bibliography: FT_1_STA-import.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = rprojroot::find_rstudio_root_file()) })
---


```{r Knitr Options, include = FALSE}
knitr::opts_chunk$set(comment = NA, message = FALSE, indent = "", error = TRUE)
```


---


# Goal of the script
This script formats the output of the resulting files from applying surface texture analysis to a sample of experimental lithics.
The script will:

1. Read in the original files  
2. Format the data  
3. Write XLSX file and save R objects ready for further analysis in R  

```{r}
dir_in  <- "analysis/raw_data"
dir_out <- "analysis/derived_data"
```

Raw data must be located in "`r paste0("./", dir_in)`".  
Formatted data will be saved in "`r paste0("./", dir_out)`".

The knit directory for this script is the project directory.


---


# Load packages
```{r}
library(grateful)
library(knitr)
library(R.utils)
library(rmarkdown)
library(tidyverse)
library(writexl)
```


---


# Read and format data
## Read in CSV file
```{r}
FT_file <- list.files(dir_in, pattern = ".*STA.*\\.csv$", full.names = TRUE) 
FT <- read.csv(FT_file, header = FALSE, na.strings = "*****", fileEncoding = 'WINDOWS-1252')
```

The following raw data file has been read: "`r FT_file`"


## Select relevant columns and rows
```{r}
FT_keep_col  <- c(4, 19:53)                    # Define columns to keep
FT_keep_rows <- which(FT[[1]] != "#")          # Define rows to keep
FT_keep      <- FT[FT_keep_rows, FT_keep_col]  # Subset rows and columns
```


## Identify results using frame numbers
```{r}
frames <- as.numeric(unlist(FT[1, FT_keep_col]))
ID <- which(frames %in% c(2, 8))
ISO <- which(frames == 10)
diriso <- which(frames %in% 11:12)
SSFA <- which(frames %in% 14:15)
furrow <- which(frames == 13)
```


## Add headers
```{r}
# Get headers from 2nd row
colnames(FT_keep) <- FT[2, FT_keep_col] %>% 
  
                     # Convert to valid names
                     make.names() %>% 
  
                     # Delete repeated periods
                     gsub("\\.+", "\\.", x = .) %>% 
  
                     # Delete periods at the end of the names
                     gsub("\\.$", "", x = .)
  
# Keep parameter name before the first period for ISO
colnames(FT_keep)[ISO] <- strsplit(names(FT_keep)[ISO], ".", fixed = TRUE) %>% 
                          sapply(`[[`, 1)

# Keep parameter name after the last period for SSFA
colnames(FT_keep)[SSFA] <- gsub("^([A-Za-z0-9]+\\.)+", "", colnames(FT_keep)[SSFA])

# Edit header non-measured point ratios
colnames(FT_keep)[ID][2] <- "NMP"

# Check results
str(FT_keep)
head(FT_keep)
```


## Extract units
```{r}
# Filter out rows which contains units
n_units <- filter(FT, V4 == "<no unit>") %>% 
          
           # Keep only unique/distinct rows of units
           distinct() %>% 
  
           # Number of unique/distinct rows of units
           nrow()

if (n_units != 1) {
  stop("The different datasets have different units")
} else {
  # Extract unit line from 3rd row for considered columns
  FT_units <- unlist(FT[3, FT_keep_col[-1]])

  # Get names associated to the units
  names(FT_units) <- colnames(FT_keep)[-1]

  # Combine into a data.frame for export
  units_table <- data.frame(variable = names(FT_units), units = FT_units)
  row.names(units_table) <- NULL
}

# Check results
units_table
```


## Split column 'Name' 
```{r}
FT_keep[c("Specimen", "State", "Location")] <- FT_keep$Name %>% 
  
                                               # Split by "_" into a matrix with 10 cols
                                               str_split_fixed("_", n = 10) %>% 
  
                                               # Keep only cols 1, 5 and 8
                                               .[, c(1, 5, 8)]
```


## Add columns for sediment type and freeze-thaw cycles
```{r}
# Load CSV file with information on samples
FT_samples_file <- list.files(dir_in, pattern = ".*samples.*\\.csv$", full.names = TRUE)
FT_samples <- read.csv(FT_samples_file)

# Merge the data.frames by chert type and chert tool
FT_keep_sed_cy <- merge(FT_keep, FT_samples, by = "Specimen")
```

The file containing the info on samples is: "`r FT_samples_file`"

Scra14 was planned for 600 cycles but the tube opened before the end of the experiment. Hence, only `r FT_samples[FT_samples$Specimen == "Scra14", "Cycles"]` cycles were conducted for that sample.


## Convert variables
```{r}
# Convert parameter variables to numeric
FT_keep_sed_cy <- type_convert(FT_keep_sed_cy)

# Convert state to factor and re-order it (1 = before, 2 = after)
FT_keep_sed_cy$State <- factor(FT_keep_sed_cy$State, levels = c("before", "after"))
```


## Set Cycles to 0 for state before
```{r}
# Replace
FT_keep_sed_cy[FT_keep_sed_cy[["State"]] == "before", "Cycles"] <- 0

# Check
FT_keep_sed_cy[c("Specimen", "State", "Cycles")]
```


## Add column for NMP categories
Here we define 3 ranges of non-measured points (NMP):  
- ≤ 10% NMP: “<10%”  
- > 10% and ≤ 17% NMP: “10-17%”  
- > 17% NMP: “>17%”

```{r}
# Create new column and fill it
FT_keep_sed_cy[FT_keep_sed_cy$NMP <= 10                           , "NMP_cat"] <- "<10%"
FT_keep_sed_cy[FT_keep_sed_cy$NMP >  10 & FT_keep_sed_cy$NMP <= 17, "NMP_cat"] <- "10-17%"
FT_keep_sed_cy[FT_keep_sed_cy$NMP >  17                           , "NMP_cat"] <- ">17%"

# Convert to ordered factor
FT_keep_sed_cy[["NMP_cat"]] <- factor(FT_keep_sed_cy[["NMP_cat"]], 
                                      levels = c("<10%", "10-17%", ">17%"), ordered = TRUE)
```

The number of height maps in each NMP category is:
```{r}
table(FT_keep_sed_cy["NMP_cat"])
```


## Add column used/unused locations
```{r}
# Define
FT_keep_sed_cy[FT_keep_sed_cy$Location %in% c("loc1", "loc2", "loc3"), "Use"] <- "used"
FT_keep_sed_cy[FT_keep_sed_cy$Location == "loc4", "Use"] <- "unused"

# Check
FT_keep_sed_cy[c("Location", "Use")]
```


## Re-order columns and add units as comment 
```{r}
FT_final <- select(FT_keep_sed_cy, Specimen, Sediment, Cycles, State, Location, Use, NMP, NMP_cat, Sq:HAsfc9)
comment(FT_final) <- FT_units
```

Type `comment(FT_final)` to check the units of the parameters.


## Check the result
```{r}
str(FT_final)
head(FT_final)
```


---


# Save data
## As XLSX
```{r}
write_xlsx(list("data" = FT_final, "units" = units_table), 
           path = paste0(dir_out, "/FT_STA_formatted-data.xlsx"))
```


## As Rbin
```{r}
saveObject(FT_final, file = paste0(dir_out, "/FT_STA_formatted-data.Rbin"))
```

Rbin files (e.g. `FT_STA_formatted-data.Rbin`) can be easily read into an R object (e.g. `rbin_data`) using the following code:
```{r, eval = FALSE}
library(R.utils)
rbin_data <- loadObject("FT_STA_formatted-data.Rbin")
```


---


# sessionInfo()

```{r}
sessionInfo()
```


---


# Cite R packages used

```{r, echo = FALSE}
pkgs_cite <- cite_packages(pkgs = "Session", omit = NULL, output = "table", include.RStudio = TRUE, 
                           out.dir = "analysis/scripts", bib.file = "FT_1_STA-import")
knitr::kable(pkgs_cite)
```


## References

