---
title: "Summary statistics on the surface texture parameters for the Freeze-thaw project"
author: "Ivan Calandra"
date: "`r format(Sys.time(), usetz = TRUE)`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cerulean
    highlight: pygments
    number_sections: true
  github_document: 
    toc: true
    toc_depth: 3
    html_preview: false
bibliography: FT_2_STA-stats.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = rprojroot::find_rstudio_root_file()) })
---


```{r Knitr Options, include = FALSE}
knitr::opts_chunk$set(comment = NA, message = FALSE, indent = "", error = TRUE)
```


---


# Goal of the script
This script computes standard descriptive statistics for each group.  
The groups are based on:

* Sediment type  
* Freeze-thaw cycles  
* Use  
* NMP_cat

It computes the following statistics:  

* sample size (n = `length`)  
* smallest value (`min`)  
* largest value (`max`)
* mean  
* median  
* standard deviation (`sd`)

```{r}
dir_in <- dir_stats <- "analysis/derived_data"
```

Raw data must be located in "`r paste0("./", dir_in)`".  
Summary statistics will be saved in "`r paste0("./", dir_stats)`".

The knit directory for this script is the project directory. This is important to specify the path for input and output.  
However, the HTML, MD and BIB files resulting from the rendering of this Rmd file will be located in the same folder as the Rmd file (i.e. "`./analysis/scripts`").

---


# Load packages
```{r}
library(doBy)
library(grateful)
library(knitr)
library(R.utils)
library(rmarkdown)
library(tidyverse)
library(writexl)
```


---


# Read in data
```{r}
FT_file <- list.files(dir_in, pattern = ".*STA.*\\.Rbin$", full.names = TRUE)
FT <- loadObject(FT_file)
```

The following data file has been read: "`r FT_file`"  

Below are its structure and first lines:
```{r}
str(FT)
head(FT)
```

---


# Summary statistics

## Create function to compute the statistics at once
```{r}
nminmaxmeanmedsd <- function(x){
    y <- x[!is.na(x)]     # Exclude NAs
    n_test <- length(y)   # Sample size (n)
    min_test <- min(y)    # Minimum
    max_test <- max(y)    # Maximum
    mean_test <- mean(y)  # Mean
    med_test <- median(y) # Median
    sd_test <- sd(y)      # Standard deviation
    out <- c(n_test, min_test, max_test, mean_test, med_test, sd_test) # Concatenate
    names(out) <- c("n", "min", "max", "mean", "median", "sd")         # Name values
    return(out)                                                        # Object to return
}
```


## Compute summary statistics
```{r}
# Compute summary statistics based on Sediment
stats_sed <- summaryBy(. ~ Sediment, data = FT, FUN = nminmaxmeanmedsd)
stats_sed[1:2]

# Compute summary statistics based on Sediment and Use
stats_sed_use <- summaryBy(. ~ Sediment + Use, data = FT, FUN = nminmaxmeanmedsd)
stats_sed_use[1:3]

# Compute summary statistics based on Sediment and NMP_cat
stats_sed_NMP <- summaryBy(. ~ Sediment + NMP_cat, data = FT, FUN = nminmaxmeanmedsd)
stats_sed_NMP[1:3]

# Compute summary statistics based on Sediment and Cycles
stats_sed_cy <- summaryBy(. ~ Sediment + Cycles, data = FT, FUN = nminmaxmeanmedsd)
stats_sed_cy[1:3]

# Compute summary statistics based on Sediment, Use and NMP_cat
stats_sed_use_NMP <- summaryBy(. ~ Sediment + Use + NMP_cat, data = FT, FUN = nminmaxmeanmedsd)
stats_sed_use_NMP[1:4]

# Compute summary statistics based on Sediment, Use and Cycles
stats_sed_use_cy <- summaryBy(. ~ Sediment + Use + Cycles, data = FT, FUN = nminmaxmeanmedsd)
stats_sed_use_cy[1:4]

# Compute summary statistics based on Sediment, Cycles and NMP_cat
stats_sed_cy_NMP <- summaryBy(. ~ Sediment + Cycles + NMP_cat, data = FT, FUN = nminmaxmeanmedsd)
stats_sed_cy_NMP[1:4]
```


## Save as XLSX
```{r}
write_xlsx(list("Sediment" = stats_sed, 
                "Sediment+Use" = stats_sed_use, 
                "Sediment+NMP" = stats_sed_NMP,
                "Sediment+Cycles" = stats_sed_cy, 
                "Sediment+Use+NMP" = stats_sed_use_NMP,
                "Sediment+Use+Cycles" = stats_sed_use_cy,
                "Sediment+Cycles+NMP" = stats_sed_cy_NMP),
           path = paste0(dir_stats, "/FT_STA-stats.xlsx"))
```


---


# sessionInfo()

```{r}
sessionInfo()
```


---


# Cite R packages used

```{r, echo = FALSE}
pkgs_cite <- cite_packages(pkgs = "Session", omit = NULL, output = "table", include.RStudio = TRUE, 
                           out.dir = "analysis/scripts", bib.file = "FT_2_STA-stats")
knitr::kable(pkgs_cite)
```


## References

