---
title: "Plots for the the Freeze-thaw project"
author: "Ivan Calandra"
date: "`r format(Sys.time(), usetz = TRUE)`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: cerulean
    highlight: pygments
    number_sections: true
  github_document: 
    toc: true
    toc_depth: 4
    html_preview: false
bibliography: FT_3_STA-plots.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = rprojroot::find_rstudio_root_file()) })
---

```{r Knitr Options, include = FALSE}
knitr::opts_chunk$set(comment = NA, message = FALSE, indent = "", error = TRUE)
```


---


# Goal of the script
The script plots all STA variables for the Freeze-thaw dataset.   

```{r}
dir_in  <- "analysis/derived_data"
dir_plots <- "analysis/plots"
```

Input Rbin data file must be located in "`r paste0("./", dir_in)`".  
Plots will be saved in "`r paste0("./", dir_plots)`".

The knit directory for this script is the project directory.


---


# Load packages
```{r}
library(doBy)
library(factoextra)
library(ggplot2)
library(grateful)
library(gridExtra)
library(knitr)
library(R.utils)
library(RColorBrewer)
library(rmarkdown)
library(tidyverse)
```


---


# Read in data
```{r}
FT_file <- list.files(dir_in, pattern = ".*STA.*\\.Rbin$", full.names = TRUE)
FT <- loadObject(FT_file)
```

The following data file has been read: "`r FT_file`"  

Below are its structure and first lines:
```{r}
str(FT)
head(FT)
```


---


# Line plots
## Filter data based on NMP_cat
```{r}
FT <- select(FT, !NMP)  %>% 
  
      # necessary only when plotting individual points and not means
      mutate(SpecLoc = paste(Specimen, Location, sep = "-"))

FT_NMP10 <- filter(FT, NMP_cat == "<10%")
FT_NMP17 <- filter(FT, NMP_cat %in% c("<10%", "10-17%"))
```

When considering only height maps with 10% NMP or less, `r nrow(FT) - nrow(FT_NMP10)` height maps are excluded and `r nrow(FT_NMP10)` are kept for further analysis.  
When considering only height maps with 17% NMP or less, `r nrow(FT) - nrow(FT_NMP17)` height maps are excluded and `r nrow(FT_NMP17)` are kept for further analysis.  


## Plotting
```{r}
# Not very readable -> plot means instead?
p <- ggplot(data = FT_NMP17, aes(x = Cycles, y = Sq, color = Sediment)) + 
     geom_point(size = 3) + 
     geom_line(linewidth = 0.5, aes(group = SpecLoc), show.legend = FALSE) +
     theme_classic() +
     
     # Set2 may be a problem for some, but not all folks with color vision impairment.
     # However, there is no colorblind safe qualitative set with 5 data classes.
     scale_color_brewer(palette = 'Set2')
print(p)
```

```{r}
# Calculate mean based on Specimen + Cycles
# Sediment is listed as factor in order to keep this column
# keep.names = TRUE is important for the matching on names in the next steps
FT_NMP17_mean <- summaryBy(.~ Specimen + Sediment + Cycles, data = FT_NMP17,
                           FUN = mean, keep.names = TRUE)

p <- ggplot(data = FT_NMP17_mean, aes(x = Cycles, y = Sq, color = Sediment)) + 
     geom_point(size = 3) + 
     geom_line(linewidth = 0.5, aes(group = Specimen), show.legend = FALSE) +
     theme_classic() +
     scale_color_brewer(palette = 'Set2')
print(p)
```


---


# PCA
## Format data
PCA will be applied to the differences between after and before the experiments, for all surface texture parameter.

The method below works only with 2 values per group, which is the case here.

```{r}
# Create function to subtract the 1st value from the 2nd value of a vector
# Returns NA if at least one value is NA
fun_minus <- function(x) {
  x[2] - x[1]
}

# Calculate the difference between after and before
                     # Exclude Cycles
FT_NMP17_pca_data <- select(FT_NMP17, !c(Cycles, NMP_cat)) %>% 
  
                     # Arrange to make sure that the difference is calculated with the correct values 
                     # in the correct order
                     arrange(Specimen, Location, State) %>% 
  
                     # Group by Specimen, Sediment and Location (= everything except State)
                     group_by(Specimen, Sediment, Location) %>%
  
                     # Apply function fun_minus to each group
                     summarize(across(where(is.numeric), fun_minus)) %>% 
  
                     # Remove rows with NA
                     na.omit() 
```


## Select surface texture parameters
The selection is based on the previous plots
```{r}
pca_params <- c("Sq", "Vmc", "Sal", "Str", 
                "Mean.density.of.furrows", "Mean.depth.of.furrows", 
                "Asfc", "HAsfc9")
```


## PCA
```{r}
pca_FT_NMP17 <- prcomp(FT_NMP17_pca_data[ , pca_params], scale. = TRUE)
```


## Plots
The library `factoextra` is not maintained anymore so there are some warnings due to issues with newer versions of ggplot2.

### Eigenvalues
```{r}
pca_FT_NMP17_eig <- fviz_eig(pca_FT_NMP17, addlabels = TRUE, ggtheme = theme_classic())
print(pca_FT_NMP17_eig)
```

### Biplots
#### Plotting function
```{r}
custom_pca_biplot <- function(dat, datpca, pc = c(1, 2), geom.pt = "point", col.pt, mean.pt = FALSE, 
                              col.pal = brewer.pal(length(unique(datpca[[col.pt]])), "Set2"), 
                              pt.size = 3, pt.shape = 19, pt.fill = "white",
                              elli = TRUE, elli.type = "convex", repel.lab = TRUE, 
                              col.variable = "black", main.title){
  
  # Define plotting
  p_out <- fviz_pca_biplot(dat, axes = pc, 
                           geom.ind = geom.pt, col.ind = datpca[[col.pt]], mean.point = mean.pt,
                           palette = col.pal, pointsize = pt.size, pointshape = pt.shape, fill.ind = pt.fill,
                           addEllipses = elli, ellipse.type = elli.type,  
                           repel = repel.lab, col.var = col.variable, title = main.title, legend.title = "")
  
  p_out <- p_out + theme(legend.position = "bottom")

  # Return plotting object
  return(p_out)
}
```

#### Biplots 
```{r}
# Define grouping variable and titles
grp_PCA <- "Sediment"

# Biplot of PC1&2
pca_FT_NMP17_12 <- custom_pca_biplot(pca_FT_NMP17, datpca = FT_NMP17_pca_data, pc = c(1, 2), col.pt = grp_PCA,
                                main.title = "PC 1&2")
print(pca_FT_NMP17_12)

# Biplot of PC3&4
pca_FT_NMP17_34 <- custom_pca_biplot(pca_FT_NMP17, datpca = FT_NMP17_pca_data, pc = c(3, 4), col.pt = grp_PCA,
                                main.title = "PC 3&4")
print(pca_FT_NMP17_34)
```

### Combine plots to save them into 1 file
```{r}
all_plots <- list(pca_FT_NMP17_eig, pca_FT_NMP17_12, pca_FT_NMP17_34)  
```

### Save plots
```{r}
#ggsave(filename = "FT_STA-PCAplots.pdf", 
#       path = dir_plots, width = 190, units = "mm", 
#       plot = marrangeGrob(all_plots, nrow = 1, ncol = 1, top = NULL))
```


---


# sessionInfo()

```{r}
sessionInfo()
```


---


# Cite R packages used

```{r, echo = FALSE}
pkgs_cite <- cite_packages(pkgs = "Session", omit = NULL, output = "table", include.RStudio = TRUE, 
                           out.dir = "analysis/scripts", bib.file = "FT_3_STA-plots")
knitr::kable(pkgs_cite)
```


## References

