---
title: "Plots for the the Freeze-thaw project"
author: "Ivan Calandra"
date: "`r format(Sys.time(), usetz = TRUE)`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
    theme: cerulean
    highlight: pygments
    number_sections: true
  github_document: 
    toc: true
    toc_depth: 4
    html_preview: false
bibliography: FT_3_STA-plots.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = rprojroot::find_rstudio_root_file()) })
---

```{r Knitr Options, include = FALSE}
knitr::opts_chunk$set(comment = NA, message = FALSE, indent = "", error = TRUE)
```


---


# Goal of the script
The script plots all STA variables for the Freeze-thaw dataset.   

```{r}
dir_in  <- "analysis/derived_data"
dir_plots <- "analysis/plots"
```

Input Rbin data file must be located in "`r paste0("./", dir_in)`".  
Plots will be saved in "`r paste0("./", dir_plots)`".

The knit directory for this script is the project directory. This is important to specify the path for input and output.  
However, the HTML, MD and BIB files resulting from the rendering of this Rmd file will be located in the same folder as the Rmd file (i.e. "`./analysis/scripts`").


---


# Load packages
```{r}
library(doBy)
library(factoextra)
library(ggplot2)
library(grateful)
library(knitr)
library(patchwork)
library(R.utils)
library(RColorBrewer)
library(rmarkdown)
library(tidyverse)
```


---


# Read in data
```{r}
FT_file <- list.files(dir_in, pattern = ".*STA.*\\.Rbin$", full.names = TRUE)
FT <- loadObject(FT_file)
```

The following data file has been read: "`r FT_file`"  

Below are its structure and first lines:
```{r}
str(FT)
head(FT)
```


---


# Format data
## Exclude NMP, and combine Specimen and Location in one column
```{r}
# Exclude NMP
FT <- select(FT, !NMP)  %>% 
  
      # Combine Specimen and Location in one column
      # Necessary to connect points when plotting individual points
      mutate(SpecLoc = paste(Specimen, Location, sep = "-")) %>%
  
      # Combine Specimen and Use in one column
      # Necessary to connect points when plotting mean points
      mutate(SpecUse = paste(Specimen, Use, sep = "-"))
```


## Create new datasets based on NMP_cat
```{r}
# Select all rows with NMP_cat ≤ 10% 
FT_NMP10 <- filter(FT, NMP_cat == "<10%")

# Select all rows with NMP_cat ≤ 10% or 10-17%
FT_NMP17 <- filter(FT, NMP_cat %in% c("<10%", "10-17%"))
```

When considering only height maps with 10% NMP or less, `r nrow(FT) - nrow(FT_NMP10)` height maps are excluded and `r nrow(FT_NMP10)` are kept for further analysis.  
When considering only height maps with 17% NMP or less, `r nrow(FT) - nrow(FT_NMP17)` height maps are excluded and `r nrow(FT_NMP17)` are kept for further analysis.  


## Calculate the mean per specimen per cycle
```{r}
# Calculate means based on Specimen + Cycles + Use
# Sediment and SpecUse are listed as factor in order to keep these columns
# keep.names = TRUE is important for the matching on names in the plots
FT_NMP10_mean <- summaryBy(.~ Specimen + Sediment + Cycles + Use + SpecUse, 
                           data = FT_NMP10, FUN = mean, keep.names = TRUE)
FT_NMP17_mean <- summaryBy(.~ Specimen + Sediment + Cycles + Use + SpecUse, 
                           data = FT_NMP17, FUN = mean, keep.names = TRUE)
```


## Add units to headers for plotting
This cannot be done before on `FT` because it create problems during the calculations of the means due to the column names with spaces and special characters (for units).  
Also, for `FT_NMP10` and `FT_NMP17`, a copy is created here because such column names are also problematic for the PCA (see below).

```{r}
# Get units from comment(FT)
table_units <- comment(FT) %>% 
               data.frame(Parameter = names(.), Unit = ., row.names = NULL) %>% 
  
               # Exclude NMP because it won't be plotted
               filter(Parameter != "NMP") %>% 
  
               # Paste parameter name and unit together in a new column
               mutate(Param_unit = paste0(Parameter, " [", Unit, "]")) 

# Remove > and < symbols
table_units$Param_unit <- gsub(">|<", "", table_units$Param_unit)

# Create copies of FT_NMP10 and FT_NMP17
FT_NMP10_units <- FT_NMP10
FT_NMP17_units <- FT_NMP17

# Adjust column names
colnames(FT_NMP10_units)[colnames(FT_NMP10_units) %in% table_units$Parameter] <- table_units$Param_unit
colnames(FT_NMP17_units)[colnames(FT_NMP17_units) %in% table_units$Parameter] <- table_units$Param_unit
colnames(FT_NMP10_mean)[colnames(FT_NMP10_mean)   %in% table_units$Parameter] <- table_units$Param_unit
colnames(FT_NMP17_mean)[colnames(FT_NMP17_mean)   %in% table_units$Parameter] <- table_units$Param_unit
```

These are the new column names for the plots:  
```{r, echo = FALSE}
cat(paste0(table_units$Param_unit, "\n"), sep = "")
```


---


# Line plots for height maps with <17% NMP
## Create list to receive the plots
```{r}
p_line_NMP17 <- vector(mode = "list", length = nrow(table_units))
names(p_line_NMP17) <- table_units$Param_unit
```


## Plots
```{r}
# Design for patchwork combination of plots (see below)
design_patch <- c(area(1, 1, 2, 3), area(3, 1, 3, 1), area(3, 2, 3, 3))

# Plot for every parameters
for (i in names(p_line_NMP17)) {
  
  # Define y-axis limits based on the range of the y-variable
  # This ensures that both plots have the same y-range
  # Not used currently
  #range_y <- range(FT_NMP17_units[[i]])
  
  # Plot of individual data points
             #  Define aesthetics 
  p_indiv <- ggplot(data = FT_NMP17_units, aes(x = Cycles, y = .data[[i]], color = Sediment, shape = Use)) + 
             
             # Add points
             geom_point(size = 3) + 
     
             # Facet plot by 'Sediment'
             facet_wrap(~ Sediment) +
    
             # Add lines to connect points with identical "SpecLoc"
             geom_line(linewidth = 0.5, aes(group = SpecLoc), show.legend = FALSE) +
    
             # Light theme
             theme_classic() +
     
             # Set2 "may be a problem for some, but not all folks with color vision impairment".
             # However, there is no colorblind safe qualitative set with 5 data classes.
             scale_color_brewer(palette = 'Set2') +
    
             # Set y-limits
             # Not used currently
             #ylim(range_y[1], range_y[2]) +
    
             # Add title to plot
             labs(title = "Individual data points") +
    
             # Position and orientation of legend
             theme(legend.direction = "horizontal", legend.title.position = "top")

  # Plot of means per sample and cycle
  p_mean <- ggplot(data = FT_NMP17_mean, aes(x = Cycles, y = .data[[i]], color = Sediment, shape = Use)) + 
            geom_point(size = 3) + 
            geom_line(linewidth = 0.5, aes(group = SpecUse), show.legend = FALSE) +
            theme_classic() +
            scale_color_brewer(palette = 'Set2') +
            #ylim(range_y[1], range_y[2]) +
            labs(title = "Mean per sample+cycle") +
            theme(legend.direction = "horizontal", legend.title.position = "top")

  # Combine both plots with patchwork
  p_line_NMP17[[i]] <- p_indiv / p_mean + guide_area() + 
                       plot_layout(guides = 'collect', design = design_patch)
}

# Print all plots
print(p_line_NMP17)
```


## Save plots
```{r}
ggsave(filename = "FT_STA-plots_NMP17.pdf", path = dir_plots, plot = p_line_NMP17, 
       width = 190, height = 200, units = "mm")
```


---


# PCA for height maps with <17% NMP
## Format data
PCA will be applied to the differences between after and before the experiments, for all surface texture parameter.

The method below works only with 2 values per group, which is the case here.

### Define function to calculate difference
```{r}
# Create function to subtract the 1st value from the 2nd value of a vector
# Returns NA if at least one value is NA
fun_minus <- function(x) {
  x[2] - x[1]
}
```

### Calculate the difference between after and before
```{r}
                     # Exclude Cycles
FT_NMP17_pca_data <- select(FT_NMP17, !c(Cycles, NMP_cat)) %>% 
  
                     # Arrange to make sure that the difference is calculated with the correct values 
                     # in the correct order
                     arrange(Specimen, Location, State) %>% 
  
                     # Group by Specimen, Sediment and Location (= everything except State)
                     group_by(Specimen, Sediment, Location) %>%
  
                     # Apply function fun_minus to each group
                     summarize(across(where(is.numeric), fun_minus)) %>% 
  
                     # Remove rows with NA
                     na.omit() 
```


## Custom plotting function
The library `factoextra` is not maintained anymore so there are some warnings due to issues with newer versions of ggplot2.

```{r}
custom_pca_biplot <- function(dat, datpca, pc = c(1, 2), geom.pt = "point", col.pt, mean.pt = FALSE, 
                              col.pal = brewer.pal(length(unique(datpca[[col.pt]])), "Set2"), 
                              pt.size = 3, pt.shape = 19, pt.fill = "white",
                              elli = TRUE, elli.type = "convex", repel.lab = TRUE, 
                              col.variable = "black", main.title){
  
  # Define plotting
  p_out <- fviz_pca_biplot(dat, axes = pc, 
                           geom.ind = geom.pt, col.ind = datpca[[col.pt]], mean.point = mean.pt,
                           palette = col.pal, pointsize = pt.size, pointshape = pt.shape, fill.ind = pt.fill,
                           addEllipses = elli, ellipse.type = elli.type,  
                           repel = repel.lab, col.var = col.variable, title = main.title, legend.title = "")
  
  p_out <- p_out + theme(legend.position = "bottom")

  # Return plotting object
  return(p_out)
}
```


## Run PCA with all parameters 
This PCA is only meant to select the most informative parameters.  
For comments on the code, see section [PCA on selected parameters](#pca-on-selected-parameters).  

```{r}
pca_FT_NMP17_all <- prcomp(FT_NMP17_pca_data[4:37], scale. = TRUE)

pca_FT_NMP17_all_eig <- fviz_eig(pca_FT_NMP17_all, addlabels = TRUE, ggtheme = theme_classic())
print(pca_FT_NMP17_all_eig)

grp_PCA <- "Sediment"
pca_FT_NMP17_all_12 <- custom_pca_biplot(pca_FT_NMP17_all, datpca = FT_NMP17_pca_data, pc = c(1, 2), 
                                         col.pt = grp_PCA, main.title = "PC 1&2")
print(pca_FT_NMP17_all_12)
pca_FT_NMP17_all_34 <- custom_pca_biplot(pca_FT_NMP17_all, datpca = FT_NMP17_pca_data, pc = c(3, 4), 
                                         col.pt = grp_PCA, main.title = "PC 3&4")
print(pca_FT_NMP17_all_34)

all_plots_17_all <- list(pca_FT_NMP17_all_eig, pca_FT_NMP17_all_12, pca_FT_NMP17_all_34)  
ggsave(filename = "FT_PCA-plots_NMP17_all-params.pdf", path = dir_plots, plot = all_plots_17_all, 
       width = 190, height = 125, units = "mm")
```


## PCA on selected parameters
### Select surface texture parameters
The following parameters are selected for the PCA:

```{r}
pca_params <- c("Sq", "Ssk", "Sv", "Sxp",
                "Vv", "Vvc", "Vm",
                "Sal", "Str", "epLsar",
                "Mean.density.of.furrows", "Mean.depth.of.furrows", "Maximum.depth.of.furrows",
                "Asfc", "HAsfc9")
cat(paste0(pca_params, "\n"), sep = "")
```

The selection was based on the [PCA with all parameters](#run-pca-with-all-parameters), trying to select parameters that contribute most to the first 4 PCs and trying to avoid parameters that correspond to the same property of the surface texture (e.g. Sa and Sq), although some surprisingly provide a different signal (e.g. Str correlating with PC1 and epLsar with PC2).  


### PCA
```{r}
pca_FT_NMP17 <- prcomp(FT_NMP17_pca_data[ , pca_params], scale. = TRUE)
```


### Plots
#### Eigenvalues
```{r}
pca_FT_NMP17_eig <- fviz_eig(pca_FT_NMP17, addlabels = TRUE, ggtheme = theme_classic())
print(pca_FT_NMP17_eig)
```

#### Biplots
```{r}
# Define grouping variable and titles
grp_PCA <- "Sediment"

# Biplot of PC1&2
pca_FT_NMP17_12 <- custom_pca_biplot(pca_FT_NMP17, datpca = FT_NMP17_pca_data, pc = c(1, 2), 
                                     col.pt = grp_PCA, main.title = "PC 1&2")
print(pca_FT_NMP17_12)

# Biplot of PC3&4
pca_FT_NMP17_34 <- custom_pca_biplot(pca_FT_NMP17, datpca = FT_NMP17_pca_data, pc = c(3, 4), 
                                     col.pt = grp_PCA, main.title = "PC 3&4")
print(pca_FT_NMP17_34)
```

#### Combine plots to save them into 1 file
```{r}
all_plots_17 <- list(pca_FT_NMP17_eig, pca_FT_NMP17_12, pca_FT_NMP17_34)  
```

#### Save plots
```{r}
ggsave(filename = "FT_PCA-plots_NMP17.pdf", path = dir_plots, plot = all_plots_17, 
       width = 190, height = 125, units = "mm")
```


---


# Line plots for height maps with <10% NMP
For comments on the code, see section [Line plots for height maps with <17% NMP](#line-plots-for-height-maps-with-17-nmp).

```{r}
p_line_NMP10 <- vector(mode = "list", length = nrow(table_units))
names(p_line_NMP10) <- table_units$Param_unit
design_patch <- c(area(1, 1, 2, 3), area(3, 1, 3, 1), area(3, 2, 3, 3))

for (i in names(p_line_NMP10)) {
  p_indiv <- ggplot(data = FT_NMP10_units, aes(x = Cycles, y = .data[[i]], color = Sediment, shape = Use)) + 
             geom_point(size = 3) + 
             facet_wrap(~ Sediment) +
             geom_line(linewidth = 0.5, aes(group = SpecLoc), show.legend = FALSE) +
             theme_classic() +
             scale_color_brewer(palette = 'Set2') +
             labs(title = "Individual data points") +
             theme(legend.direction = "horizontal", legend.title.position = "top")
  p_mean <- ggplot(data = FT_NMP10_mean, aes(x = Cycles, y = .data[[i]], color = Sediment, shape = Use)) + 
            geom_point(size = 3) + 
            geom_line(linewidth = 0.5, aes(group = SpecUse), show.legend = FALSE) +
            theme_classic() +
            scale_color_brewer(palette = 'Set2') +
            labs(title = "Mean per sample+ cycle") +
            theme(legend.direction = "horizontal", legend.title.position = "top")
  p_line_NMP10[[i]] <- p_indiv / p_mean + guide_area() + 
                       plot_layout(guides = 'collect', design = design_patch)
}

print(p_line_NMP10)
ggsave(filename = "FT_STA-plots_NMP10.pdf", path = dir_plots, plot = p_line_NMP10, 
       width = 190, height = 200, units = "mm")
```


---


# sessionInfo()

```{r}
sessionInfo()
```


---


# Cite R packages used

```{r, echo = FALSE}
pkgs_cite <- cite_packages(pkgs = "Session", omit = NULL, output = "table", include.RStudio = TRUE, 
                           out.dir = "analysis/scripts", bib.file = "FT_3_STA-plots")
knitr::kable(pkgs_cite)
```


## References

